Assessing the Impact of MDE and Code Generation on the Sustainability of Scientific Computing Software: A Research Proposal

### 2019-09-16 ResearchProposal.tex
Additional edits to research proposal - fleshing out introductory sections

Without dramatic intervention, our collective confidence in Scientific Computing
Software (SCS) is due for a catastrophic collapse.  We are increasingly trusting
ever more complex and ambitious computations, but our software foundations are
built on sand. Areas of concern include nuclear safety analysis and
computational medicine. Successfully building such software requires
communication between software developers and experts from multiple
domains. Collaboration is difficult at the best of times, and is made worse
because developers favour handcrafted solutions over adapting software
engineering processes, methods and tools \citep{FaulkEtAl2009}. Handcrafted
solutions do not account for the inevitable changes in requirements, design and
implementation. The twin challenges of changing requirements and inadequate
documentation conspire to make computational results notoriously difficult to
reproduce, especially in the important endeavour of one researcher independently
replicating the results of another \citep{Smith2018}.  Inadequate traceability
within documentation means re­certification of previously certified engineering
software is as expensive and time consuming as the original certification
exercise, because of a lack of explicit, effort reducing, traceability
information.

Documentation can address the confidence problem mentioned above.  For software
in general, documentation, written before and during development, provides many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements.  For SCS in particular, developers
do not always value documentation, but documentation has been shown to help
improve SCS.  For instance, \citet{SmithEtAl2015-SS-TR} shows that the quality
of statistical software for psychology is generally improved when developed
using the structured CRAN (Comprehensive R Archive Network) process and tools,
versus an ad hoc process.  \citet{SmithAndKoothoor2016} highlights the value of
proper documentation by redeveloping nuclear safety analysis software.  Twenty
seven (27) worrisome documentation problems were found, including
incompleteness, ambiguity, inconsistency, verifiability, modifiability,
traceability and a lack of abstraction.  A redevelopment experiment with five
existing projects \citep{SmithJegatheesanAndKelly2016} enabled the code owners
(as ascertained through interviews) to clearly see the value of documentation.
However, mirroring other studies \citep{CarverEtAl2007}, the code owners felt
documentation takes too much time.

One promising approach to gain the benefits of documentation, without the often
infeasible time commitment, is to use a Model Driven Engineering (MDE), where
MDE is an engineering approach that exploits structured representations
(metamodels for abstract syntax, and constraints to capture static semantics) to
automate repetitive, tedious and error prone tasks.  In this research proposal,
we will use MDE as a short-hand to include model transformation, Domain Specific
Languages (DSLs) and code generation.  Several current papers point to DSLs and
code/document generation as a transformative technology for documentation,
design and verification of SCS \citep{JohansonAndHasselbring2018, Smith2018}.
The benefit of an MDE approach is to dramatically reduce the cost of developing
reliable, reproducible and re-certifiable software by: i) creating
infrastructure for a knowledge base of scientific and computing models; and, ii)
writing explicit ``recipes'' that weave together this knowledge to generate
theoretical models, design documents, code, test cases and build scripts. The
generator can render to multiple languages, such as html, LaTeX, RTF for
documentation and Python, Java, C, and Matlab, for code. One source of
knowledge, with rules for transformation, means completeness, consistency and
traceability can be achieved by construction. Moreover, these qualities can be
maintained as requirements are modified, design decisions are changed,
documentation standards are varied, and software is re­certified.  The knowledge
will be built up incrementally.  New projects will reuse existing knowledge and
expand as necessary.

Scientists are awakening to the critical importance of SCS sustainability, as
evidenced by the success of Software Carpentry and the Software
Sustainability Institute, but the current solutions are not bold enough.
Although valuable, we need more than studies of developers, training sessions
and project support tools.  \emph{We need transformative change (enabled by
  creating new tools for knowledge capture and document/code generation) where
  documentation, design and verification are expected for all SCS.}

- background and context

- problem statement and research question
- research question: What ideas (if any) from Drasil and Drasil procedures are perceived
by graduate students and end user developers with improving software quality and
productivity - for long lived (safety related?) projects

roadmap of document

\section{Background and Literature Review}

roadmap of this section.  Some overview of the topics themselves, but an
emphasis on how they appear in the SCS context.

\subsection{Software Qualities of Interest}

For instance,
a rational design process can provide quality improvements, and Quality
Assurance (QA), as shown in Table~\ref{Tbl_Qualities}.  Moreover, documentation,
written before and during development, can provide many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements.

- qualities of interest, add sustainability

@@

@@

% \item [\textbf{Reliability}] is a critical quality for scientific software,
%   since the results of computations are meaningless, if they are not dependable.
%   Reliability is closely tied to verifiability, since the key quality to verify
%   is reliability, while the act of verification itself improves reliability.

% \item [\textbf{Performance}] considerations can make certification challenging, since QA
%   becomes more difficult for more complex code.  However,
%   as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
%   reducing the number of options in general purpose code, is not always a
%   practical option.

@@

\item [\textbf{Reproducibility}] is a required component of the scientific
  method~\cite{Davison2012}.  Although QA has, ``a bad name among
  creative scientists and engineers''~\cite[p.~352]{Roache1998}, the community
  need to recognize that participating in QA management also improves
  reproducibility.  Reproducibility, like QA, benefits from a consistent and
  repeatable computing environment, version control and separating code from
  configuration/parameters~\cite{Davison2012}.
\end{description}

\subsection{Desirable Qualities of Documentation}

To address the challenges for adequate documentation for QA and certification
efforts for SCS, the qualities described below need to be a priority.  All but
the final quality listed (abstraction), are adapted from the IEEE recommended
practise for producing good software requirements~\cite{IEEE1998}.

\begin{itemize}

\item \emph{Completeness}: Documentation is said to be complete when all the
  requirements of the software are detailed. That is, each goal, functionality,
  attribute, design constraint, value, data, model, symbol, term (with its unit
  of measurement if applicable), abbreviation, acronym, assumption and
  performance requirement of the software is defined.  The software's response
  to all classes of inputs, both valid and invalid and for both desired and
  undesired events, also needs to be specified.

\item \emph{Consistency}: Documentation is said to be consistent when no subset
  of individual statements are in conflict with each other. That is, a
  specification of an item made at one place in the document should not
  contradict the specification of the same item at another location.

\item \emph{Modifiability}: The documentation should be developed in such a way
  that it is easily modifiable so that likely future changes do not destroy the
  structure of the document. Also it should be easy to reflect the change,
  wherever needed in the document to maintain consistency, traceability and
  completeness. For documentation to be modifiable, its format must be
  structured in a way that repetition is avoided and cross-referencing is
  employed.

\item \emph{Traceability}: Documentation should be traceable, as this
  facilitates maintenance and review. If a change is made to the design or code
  of the software, then all the documentation relating to those segments have to
  be modified.  This property is also important for recertification.

\item \emph{Unambiguity}: Documentation is said to be unambiguous only when
  every requirement's specification has a unique interpretation.  The
  documentation should be unambiguous to all audiences, including developers,
  users and reviewers.

\item \emph{Correctness}: There is no direct tool or method for measuring
  correctness. One way of building confidence in correctness is by reviewing to
  ensure that each requirement stated is one that the stakeholders and experts
  desire.  By maintaining traceability, consistency and unambiguity, we can
  reduce the occurrence of errors and make the goal of reviewing for correctness
  easier.

\item \emph{Verifiability}: Every requirement in the documentation must be the
  one fulfilled by the implemented software. Therefore all the requirements
  should be clear, unambiguous and testable, so that a person or a machine can
  verify whether the software product meets the requirements.

\item \emph{Abstract}: Documented requirements are said to be abstract if they
  state what the software must do and the properties it must possess, but do not
  speak about how these are to be achieved. For example, a requirement can
  specify that an Ordinary Differential Equation (ODE) must be solved, but it
  should not mention that Euler's method should be used to solve the ODE. How to
  accomplish the requirement is a design decision, which is documented during
  the design phase.

\end{itemize}

\subsection{Current State of Practice for SCS Development}

@@

\subsection{Documentation for SCS}

@@

\subsection{MDE and Code Generation}

DSLs and code/document generation provide a transformative technology for
documentation, design and verification \citep{JohansonAndHasselbring2018,
  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
science not software \citep{Kelly2007}.  A generative approach removes the
maintenance nightmare of documentation duplicates and near duplicates
\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
transformed as needed.  Code generation has previously been applied to improve
SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.
%   For instance,
% ATLAS (Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} 
%and
%Blitz++ \citep{Veldhuizen1998} 
% produces efficient and portable linear algebra
% software.  
% Spiral \citep{Pueschel2001} uses software/hardware generation for
% digital signal processing.  
\citet{Carette2008} shows how to generate a family
of efficient, type-safe Gaussian elimination algorithms.  FEniCS (Finite Element
and Computational Software) \citep{LoggEtAl2012} uses code generation when
solving differential equations.  Unlike previous work on SCS code generation,
Drasil will focus on generating all software artifacts (requirements, design
etc.), not just code.

@@

\subsection{Empirical Methods for Software Engineering}

@@

\section{Research Design and Methodologies}

@@

\subsection{State of the Practice}

@@

\subsection{Quantum Chemistry Software}

@@

\subsection{Impact of MDE on SCS}

@@

\subsection{Fitting Drasil into the Scientific Software Development Process}

@@

\subsection{etc.}

@@

\subsection{Impact of Assurance Cases for Building Confidence}

@@

\subsection{Impact of Computational Variability Testing}

@@

\section{Implications and Contributions to Knowledge}

@@

\section{Research Schedule}

>>>>

Without dramatic intervention, our collective confidence in Scientific Computing
Software (SCS) is due for a catastrophic collapse.  We are increasingly trusting
ever more complex and ambitious computations, but our software foundations are
built on sand. Areas of concern include nuclear safety analysis and
computational medicine. Successfully building such software requires
communication between software developers and experts from multiple
domains. Collaboration is difficult at the best of times, and is made worse
because developers favour handcrafted solutions over adapting software
engineering processes, methods and tools \citep{FaulkEtAl2009}. Handcrafted
solutions do not account for the inevitable changes in requirements, design and
implementation. The twin challenges of changing requirements and inadequate
documentation conspire to make computational results notoriously difficult to
reproduce, especially in the important endeavour of one researcher independently
replicating the results of another \citep{Smith2018}.  Inadequate traceability
within documentation means re­certification of previously certified engineering
software is as expensive and time consuming as the original certification
exercise, because of a lack of explicit, effort reducing, traceability
information.  The existing challenges for modification, maintenance and
extension point to problems with the sustainability of SCS.

Documentation can improve software sustainability.  For software
in general, documentation, written before and during development, provides many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements. For SCS in particular,
documentation provides benefits.  For instance, \citet{SmithEtAl2015-SS-TR}
shows that the quality of statistical software for psychology is generally
improved when developed using the structured CRAN (Comprehensive R Archive
Network) process and tools, versus an ad hoc process.
\citet{SmithAndKoothoor2016} highlights the value of proper documentation by
redeveloping nuclear safety analysis software.  Twenty seven (27) worrisome
documentation problems were found, including incompleteness, ambiguity,
inconsistency, verifiability, modifiability, traceability and a lack of
abstraction.  A redevelopment experiment with five existing projects
\citep{SmithJegatheesanAndKelly2016} enabled the code owners (as ascertained
through interviews) to clearly see the value of documentation.  However,
mirroring other studies \citep{CarverEtAl2007}, the code owners felt
documentation takes too much time.  Due to the time commitment and resource
limitation, documentation is rarely emphasized in SCS.

One promising approach to gain the benefits of documentation, but significantly
reduce the time commitment, is to use a Model Driven Engineering (MDE).  MDE is
an engineering approach that exploits structured representations (metamodels for
abstract syntax, and constraints to capture static semantics) to automate
repetitive, tedious and error prone tasks.  In this research proposal, we will
use MDE as a short-hand to include knowledge models, model transformation,
Domain Specific Languages (DSLs) and code generation.  Several current papers
point to DSLs and code/document generation as a transformative technology for
documentation, design and verification of SCS \citep{JohansonAndHasselbring2018,
  Smith2018}.  The benefit of an MDE approach is to dramatically reduce the cost
of developing reliable, reproducible and re-certifiable software.

One example of applying MDE to SCS is Drasil, which is introduced in
\citet{SzymczakEtAl2016} and available at
\href{https://github.com/JacquesCarette/Drasil}
{https://github.com/JacquesCarette/Drasil}.  Drasil is implemented as an open
source set of Domain Specific Languages (DSLs).  The Drasil approach involves:
i) creating infrastructure for a knowledge base of scientific and computing
models; and, ii) writing explicit ``recipes'' that weave together this knowledge
to generate theoretical models, design documents, code, test cases and build
scripts. The generator can render to multiple languages, such as html, LaTeX,
RTF for documentation and Python, Java, C, and Matlab, for code. One source of
knowledge, with rules for transformation, means completeness, consistency and
traceability can be achieved by construction. Moreover, these qualities can be
maintained as requirements are modified, design decisions are changed,
documentation standards are varied, and software is re­certified.  Although
creating the knowledge base is time consuming, the knowledge can be built up
incrementally.  New projects will reuse existing knowledge and expand as
necessary.

As described above, sustainability is a concern for SCS, and MDE holds promise
for addressing this concern.  However, a transformative change in the
development of SCS requires more than promise.  Empirical evidence is necessary
to demonstrate that an MDE process can improve the sustainability of SCS.  This
leads to the long-term objective for this research proposal: \emph{assess and
  measure the impact of MDE on the sustainability of SCS.}

The scope of SCS software for this proposal includes long-lived software, as
implied by the sustainability objective.  To gain the full benefits of
generating documentation, the scope also includes software that interests
multiple stakeholders, with different interests and backgrounds.  Finally, the
scope of this proposal emphasizes safety related software, such as software for
nuclear safety analysis, medical imaging and computational medicine.  Safety
related software is prioritized because documentation is often expected as part
of a certification exercise.  Although this study is not specifically restricted
to open-source software, for practical reasons, commercial software will not be
strongly emphasized.

Given the large number of potential MDE processes, techniques and technologies,
together with the wide variety in purpose, scope and context for SCS, the
long-term objective of measuring the impact of MDE is decomposed into four
short-term objectives.  For each objective, a reference is given to the
corresponding section in this document that provides the details.

\begin{enumerate}
\item Assess the current state of software development practice for SCS.  How is
  the software developed?  Are there existing development methods that lead to
  higher software sustainability? (Section~\ref{SecStateOfThePractice})
\item Assess the impact of an MDE process on end user developed SCS with respect
  to developer productivity and software sustainability.  
  (Section~\ref{SecImpactMDE})
\item Assess the impact on safety when the MDE generated documentation targets a
  safety assurance case.  An assurance case is an explicit structured argument
  pertaining to specific properties, such as trustworthiness.  The evidence for
  the argument will come from the traceability between artifacts generated by
  model transformations, expert reviews, test cases,
  etc. (Section~\ref{SecAssuranceCase})
\item Assess the impact on software quality when using MDE to facilitate
  Computational Variability Testing (CVT).  CVT uses code generation to build
  confidence in the generated code in an analogous way to the use of grid
  refinement studies.  Grid refinement looks at how the solution changes by
  varying the run-time parameter of grid density and comparing the results to
  theoretical expectations.  CVT, on the other hand, can generate code to
  ``refine'' build time parameters, such as order of interpolation, or degree of
  implicitness.  These parameters can be systematically varied and the results
  compared against the expected trend.  (Section~\ref{SecCVT})
\end{enumerate}

Before presenting the details of the short-term objectives
(Section~\ref{SecMethodologies}), the necessary background information is
provided, along with a literature review (Section~\ref{SecBackgroundLitReview}).
The literature review covers software quality (Sections~\ref{SecSoftwareQuality}
and~\ref{SecDesirableQs}), the literature on the current state of the practice
for SCS development (Section~\ref{SecCurrStateOfThePractice}), potential target
documentation artifacts for SCS (Section~\ref{SecDocumentation}), MDE and code
generation (Section~\ref{SecMDE}) and empirical methods for software engineering
(Section~\ref{SecEmpirical}).  This document ends with a summary of the
potential contributions to knowledge from this study
(Section~\ref{SecContribKnowledge}) and a research schedule
(Section~\ref{SecSchedule}).

\section{Background and Literature Review} \label{SecBackgroundLitReview}

To assess the impact of MDE on SCS quality, we need a clear definition of what
we mean by quality.  Section~\ref{SecSoftwareQuality} shows how the concept of
quality is decomposed into a set of separate qualities.  This set of qualities
can be applied to the software artifacts (documentation, test cases, etc) and to
the software development process itself.  Several of the qualities from the list
in Section~\ref{SecSoftwareQuality} cannot be measured directly, such as
maintainability.  Therefore, Section~\ref{SecDesirableQs} introduces measurable
documentation qualities that are believed to contribute to software quality.  In
some cases it may be necessary to use the qualities in
Section~\ref{SecDesirableQs} to indirectly measure qualities listed in
Section~\ref{SecSoftwareQuality}.

focuses on presents it, roadmap of this
section.  Some overview of the topics themselves, but an emphasis on how they
appear in the SCS context.

\subsection{Software Qualities of Interest} \label{SecSoftwareQuality}

Our analysis is centred around a set of software qualities.  Quality is not
considered as a single measure, but a collection of different qualities, often
called ``ilities.''  These qualities highlight the desirable nonfunctional
properties for software artifacts, which include both documentation and
code. Some qualities, such as visibility and productivity, apply to the process
used for developing the software. The following list of qualities is based on
\cite{GhezziEtAl2003}. To the list from \cite{GhezziEtAl2003}, we have added
three qualities important for SC: installability, reproducibility and
sustainability.

@@

\item [\textbf{Installability}] A measure of the ease of installation.

\item [\textbf{Correctness}] Software is correct if it matches its specification.

@@

\item [\textbf{Reliability}] is a critical quality for scientific software,
  since the results of computations are meaningless, if they are not dependable.
  Reliability is closely tied to verifiability, since the key quality to verify
  is reliability, while the act of verification itself improves reliability.

\item [\textbf{Performance}] considerations can make certification challenging, since QA
  becomes more difficult for more complex code.  However,
  as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
  reducing the number of options in general purpose code, is not always a
  practical option.

@@

\item [\textbf{Reproducibility}] is a required component of the scientific
  method~\cite{Davison2012}.  Although QA has, ``a bad name among
  creative scientists and engineers''~\cite[p.~352]{Roache1998}, the community
  need to recognize that participating in QA management also improves
  reproducibility.  Reproducibility, like QA, benefits from a consistent and
  repeatable computing environment, version control and separating code from
  configuration/parameters~\cite{Davison2012}.

\item [\textbf{Productivity}] [Fill in]

\item [\textbf{Sustainability}] A combination of other qualities.

\end{description}

\subsection{Desirable Qualities of Documentation} \label{SecDesirableQs}

To achieve the qualities listed in Section~\ref{SecSoftwareQuality}, the
documentation should achieve the qualities listed in this section.  All but the
final quality listed (abstraction), are adapted from the IEEE recommended
practise for producing good software requirements~\cite{IEEE1998}.

\begin{description}

\item [\textbf{Completeness}] Documentation is said to be complete when all the
  requirements of the software are detailed. That is, each goal, functionality,
  attribute, design constraint, value, data, model, symbol, term (with its unit
  of measurement if applicable), abbreviation, acronym, assumption and
  performance requirement of the software is defined.  The software's response
  to all classes of inputs, both valid and invalid and for both desired and
  undesired events, also needs to be specified.

\item [\textbf{Consistency}] Documentation is said to be consistent when no subset
  of individual statements are in conflict with each other. That is, a
  specification of an item made at one place in the document should not
  contradict the specification of the same item at another location.

\item [\textbf{Modifiability}] The documentation should be developed in such a way
  that it is easily modifiable so that likely future changes do not destroy the
  structure of the document. Also it should be easy to reflect the change,
  wherever needed in the document to maintain consistency, traceability and
  completeness. For documentation to be modifiable, its format must be
  structured in a way that repetition is avoided and cross-referencing is
  employed.

\item [\textbf{Traceability}] Documentation should be traceable, as this
  facilitates maintenance and review. If a change is made to the design or code
  of the software, then all the documentation relating to those segments have to
  be modified.  This property is also important for recertification.

\item [\textbf{Unambiguity}] Documentation is said to be unambiguous only when
  every requirement's specification has a unique interpretation.  The
  documentation should be unambiguous to all audiences, including developers,
  users and reviewers.

\item [\textbf{Correctness}] There is no direct tool or method for measuring
  correctness. One way of building confidence in correctness is by reviewing to
  ensure that each requirement stated is one that the stakeholders and experts
  desire.  By maintaining traceability, consistency and unambiguity, we can
  reduce the occurrence of errors and make the goal of reviewing for correctness
  easier.

\item [\textbf{Verifiability}] Every requirement in the documentation must be the
  one fulfilled by the implemented software. Therefore all the requirements
  should be clear, unambiguous and testable, so that a person or a machine can
  verify whether the software product meets the requirements.

\item [\textbf{Abstract}] Documented requirements are said to be abstract if they
  state what the software must do and the properties it must possess, but do not
  speak about how these are to be achieved. For example, a requirement can
  specify that an Ordinary Differential Equation (ODE) must be solved, but it
  should not mention that Euler's method should be used to solve the ODE. How to
  accomplish the requirement is a design decision, which is documented during
  the design phase.

\end{description}

\subsection{Literature Review on Current State of the Practice for SCS
  Development} \label{SecCurrStateOfThePractice}

@@

\subsection{Documentation for SCS} \label{SecDocumentation}

@@

\subsection{MDE and Code Generation} \label{SecMDE}

DSLs and code/document generation provide a transformative technology for
documentation, design and verification \citep{JohansonAndHasselbring2018,
  Smith2018}.  DSLs allow scientists their preferred approach of focusing on
science not software \citep{Kelly2007}.  A generative approach removes the
maintenance nightmare of documentation duplicates and near duplicates
\citep{LucivEtAl2018}, since knowledge is only captured once and automatically
transformed as needed.  Code generation has previously been applied to improve
SCS \citep{WhaleyEtAl2001, Veldhuizen1998, Pueschel2001}.  For instance, ATLAS
(Automatically Tuned Linear Algebra Software) \citep{WhaleyEtAl2001} and Blitz++
\citep{Veldhuizen1998} produces efficient and portable linear algebra software.
Spiral \citep{Pueschel2001} uses software/hardware generation for digital signal
processing.  \citet{Carette2008} shows how to generate a family of efficient,
type-safe Gaussian elimination algorithms.  FEniCS (Finite Element and
Computational Software) \citep{LoggEtAl2012} uses code generation when solving
differential equations.  Unlike previous work on SCS code generation, Drasil
will focus on generating all software artifacts (requirements, design etc.), not
just code.

@@

\subsection{Empirical Methods for Software Engineering} \label{SecEmpirical}

@@

\section{Research Design and Methodologies} \label{SecMethodologies}

@@

\subsection{State of the Practice} \label{SecStateOfThePractice}

@@

\subsubsection{Quantum Chemistry Software}

@@

\subsection{Impact of MDE on SCS} \label{SecImpactMDE}

@@

Research Question: What tools and methods from MDE are perceived by end user developers to
  improve their productivity and the sustainability of their software.

\subsubsection{Fitting Drasil into the Scientific Software Development Process} \label{Sec}

@@

\subsubsection{etc.}

@@

\subsection{Impact of Assurance Cases for Building Confidence} \label{SecAssuranceCase}

@@

\subsection{Impact of Computational Variability Testing} \label{SecCVT}

@@

\section{Implications and Contributions to Knowledge} \label{SecContribKnowledge}

@@

\section{Research Schedule} \label{SecSchedule}
